---
title: "Extended R script of Chapt.6  __Nonlinear ME Models__"
author: "Pinheiro, Bates (2000) and Maechler (2018, s. below)"
date: "May 18, 2020"
output:
  html_document:
    toc: TRUE
    keep_md: TRUE
  pdf_document:
    toc: TRUE
---

From **The Book**
#  Pinheiro, J.C., and Bates, D.M. (2000)
#  *Mixed-Effects Models in S and S-PLUS*, Springer.

- Copyright (2000) Jose Pinheiro and Douglas Bates (Book!!)
- Copyright (2018) Martin Maechler (transl. to `Rmd`; additional plots, etc)

# Ch. 6 Nonlinear ME Models: Basic Concepts and Motivating Examples

```{r, setup}
library(nlme)
options(width = 75  # MM: width 65 is too narrow
      , digits = 5
      , contrasts = c(unordered = "contr.helmert",
                        ordered = "contr.poly") )
if(!dev.interactive(TRUE)) pdf(file = 'ch06_plot.pdf')
```{r, setup2, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE
  , comment = NA  # MM: R output is *not* comment!   was "#>"
  , prompt = TRUE # MM: yes, should look like R session
  , fig.width = 9 # MM: default 7 (too narrow)
)
```

## 6.2 Indomethicin Kinetics

```{r, Indometh}
plot(Indometh)
## 1. a "global" model fit to all 6 subjects :
fm1Indom.nls <- nls(conc ~ SSbiexp(time, A1, lrc1, A2, lrc2),
                    data = Indometh)
summary(fm1Indom.nls)
plot(fm1Indom.nls, Subject ~ resid(.), abline = 0,   # Fig 6.4
     main = "nls() global fit")

## 2. Subject-specific model: Each fit *separately* all 6 subjects :
(fm1Indom.lis <- nlsList(conc ~ SSbiexp(time, A1, lrc1, A2, lrc2),
                         data = Indometh))
plot(fm1Indom.lis, Subject ~ resid(.), abline = 0,
     main = "nlsList(): individual fits | Subject") # Fig 6.5
plot(intervals(fm1Indom.lis), main = "intervals( nlsList(*) )") # Fig. 6.6

## 3. Mixed-Effects Model: "Global" model + subject-specific *random* effect
(fm1Indom.nlme <- nlme(fm1Indom.lis,
                      random = pdDiag(A1 + lrc1 + A2 + lrc2 ~ 1),
                      control = list(tolerance = 0.0001)))
                                        # sigma_{lrc2} ~= 0  ==> drop it :
## not in book, compare with 6.5:
if(dev.interactive()) dev.new()
plot(fm1Indom.nlme, Subject ~ resid(.), abline = 0, col="gray",
     main = "nlme(): global + RE | Subject")

fm1Indom.nlme # sigma_{lrc2} is almost 0 ==> set to 0
fm2Indom.nlme <- update(fm1Indom.nlme,
                        random = pdDiag(A1 + lrc1 + A2 ~ 1))
anova(fm1Indom.nlme, fm2Indom.nlme)# almost same fit in smaller model: keep it

# Consider correlations between RE's
(fm3Indom.nlme <- update(fm2Indom.nlme, random = A1+lrc1+A2 ~ 1))
                  # warning "did not converge" .. but only in iter. 1

## try to "obey" and increase msMaxIter: --> next limit
update(fm2Indom.nlme, random = A1+lrc1+A2 ~ 1,
                         control = nlmeControl(msMaxIter = 500))



# only keep correlation in "first block" (A1, lrc1);  set ran.eff.(lrc2) to 0

fm4Indom.nlme <-
    update(fm3Indom.nlme,
           random = pdBlocked(list(A1 + lrc1 ~ 1, A2 ~ 1)))
fm4Indom.nlme
anova(fm3Indom.nlme, fm4Indom.nlme) # => "equivalent"; '4' is smaller
anova(fm2Indom.nlme, fm4Indom.nlme) # => '4' >> '2' clearly
plot(fm4Indom.nlme, id = 0.05, adj = -1)# Tukey-Anscombe plot (resid ~ fitted)
                                        # slightly differing from book's  Fig. 6.7
qqnorm(fm4Indom.nlme) # help( qqnorm.lme )
qqnorm(fm4Indom.nlme, id = 0.05, adj = -1) # --> again: it is 'Subject 2'
qqnorm(fm4Indom.nlme, ~ resid(., type = "p") | Subject, abline = c(0, 1))

summary(fm4Indom.nlme) # (cf. book , p. 286-7) -- R: larger likelihood
## compare Fixed eff. w/ the global model FE's (and Std.Err. !):
coef(summary(fm1Indom.nls))

## Nice plot :
plot(augPred(fm4Indom.nlme, level = 0:1, grid=TRUE)) # help( plot.augPred )
```

## 6.3 Growth of Soybean Plants

```{r, Soybean}
head(Soybean)
summary(Soybean)
plot(Soybean, outer = ~ Year * Variety)
## nlsList() :  an nls() model for each year:Variety combination
(fm1Soy.lis <- nlsList(weight ~ SSlogis(Time, Asym, xmid, scal),
                       data = Soybean))
## -- warnings: in R, 2 cases did not converge (1989 P5 &  1989 P8)

## Starting nlme() from nlsList() fit (!) :
(fm1Soy.nlme <- nlme(fm1Soy.lis)) # takes 2-3 sec
plot(fm1Soy.nlme, id = 0.01, adj = -.1)# Tukey-Anscombe plot (resid ~ fitted)
                                        # --> not good!
## nlme allows *modeling*  weights = varPower() :
fm2Soy.nlme <- update(fm1Soy.nlme, weights = varPower())
## convergence warnings
plot(fm2Soy.nlme, id = 0.01, adj = -.1)# TA plot -- much better

## the three  non-linear regr. coefficients are subject specific - random  (!)
plot(ranef(fm2Soy.nlme, augFrame = TRUE),
     form = ~ Year * Variety, layout = c(3,1))
## ==> NB: the three logistic parameters are very very highly correlated.
round(cor(ranef(fm2Soy.nlme)), 5)
## less extreme from here:
summary(fm2Soy.nlme)## => Ran.Eff: corr(b_Asym, b_xmid) ~= -1
                    ##  & Fix.Eff: corr(xmid, scal)     ~= 0.77
##==> Want a better model

(soyFix <- fixef(fm2Soy.nlme)) # to use for 'start' below:
options(contrasts = c("contr.treatment", "contr.poly")) # for easier interpretation
(fm3Soy.nlme <-
 update(fm2Soy.nlme,
        fixed = Asym + xmid + scal ~ Year,
        start = c(soyFix[1], 0, 0, soyFix[2], 0, 0, soyFix[3], 0, 0)))
                                        # NB:  power^ = 0.95 ~= 1
anova(fm3Soy.nlme)
```
The following line is not in the book but needed to fit the model

```{r }
fm4Soy.nlme <-
    nlme(weight ~ SSlogis(Time, Asym, xmid, scal),
         data = Soybean,
         fixed = list(Asym ~ Year*Variety, xmid ~ Year + Variety, scal ~ Year),
         random = Asym ~ 1,
         start = c(17, 0, 0, 0, 0, 0, 52, 0, 0, 0, 7.5, 0, 0),
         weights = varPower(0.95), control = list(verbose = TRUE))
```
   NB (nlme-FIXME): update()  does not work for the fixed argument when fixed is a list

```{r }
## p. 293-4 :
summary(fm4Soy.nlme)
plot(augPred(fm4Soy.nlme))# Fig 6.14, p. 295  <<<<<<<<<<< Final Fit (+ Data)
```

## 6.4 Clinical Study of Phenobarbital Kinetics

```{r, fm1Pheno.nlme}
(fm1Pheno.nlme <-
 nlme(conc ~ phenoModel(Subject, time, dose, lCl, lV),
      data = Phenobarb, fixed = lCl + lV ~ 1,
      random = pdDiag(lCl + lV ~ 1), start = c(-5, 0),
      na.action = NULL, naPattern = ~ !is.na(conc)))
fm1Pheno.ranef <- ranef(fm1Pheno.nlme, augFrame = TRUE)
```

FIXME: these plots encounter difficulties

```{r }
plot(fm1Pheno.ranef, form = lCl ~ Wt + ApgarInd)
plot(fm1Pheno.ranef, form = lV ~ Wt + ApgarInd)
options(contrasts = c("contr.treatment", "contr.poly"))
## This fit just ping-pongs
##fm2Pheno.nlme <-
##    update(fm1Pheno.nlme,
##           fixed = list(lCl ~ Wt, lV ~ Wt + ApgarInd),
##           start = c(-5.0935, 0, 0.34259, 0, 0),
##           control = list(pnlsTol = 1e-4, maxIter = 500,
##           msVerbose = TRUE))
##summary(fm2Pheno.nlme)
##fm3Pheno.nlme <-
##    update(fm2Pheno.nlme,
##           fixed = lCl + lV ~ Wt,
##           start = fixef(fm2Pheno.nlme)[-5])
##plot(fm3Pheno.nlme, conc ~ fitted(.), abline = c(0,1))
```

cleanup

```{r }
proc.time()
```

