---
title: "Extended R script of Chapt.8  __Fitting Nonlinear ME Models__"
author: "Pinheiro, Bates (2000) and Maechler (2018-9, s. below)"
date: "May 27, 2019"
output:
  html_document:
    toc: TRUE
    keep_md: TRUE
  pdf_document:
    toc: TRUE
---

From **The Book**
#  Pinheiro, J.C., and Bates, D.M. (2000)
#  *Mixed-Effects Models in S and S-PLUS*, Springer.

- Copyright (2000) Jose Pinheiro and Douglas Bates (Book!!)
- Copyright (2018) Martin Maechler (transl. to `Rmd`; additional plots, etc)

# Ch. 8    Fitting Nonlinear Mixed-Effects Models

## 8.1 Fitting Nonlinear Models in S with nls and nlsList

```{r, setup, echo=FALSE}
library(nlme)
library(lattice)
options(width = 75, digits = 5) # MM: 65 is too narrow
options(contrasts = c(unordered = "contr.helmert", ordered = "contr.poly"))
if(!dev.interactive(TRUE)) pdf(file = 'ch08_plot.pdf')
```
```{r, setup2, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE
  , comment = NA   # MM: R output is *not* comment!   was "#>"
  , prompt = TRUE  # MM: yes, should look like R session
  , fig.width = 9  # MM: default is 7
)
```

###  8.1.1  Using the `nls()` Function


```{r, Orange-logist}
## outer = ~1 is used to display all five curves in one panel
plot(Orange, outer = ~1)
logist <-
   function(x, Asym, xmid, scal) Asym/(1 + exp(-(x - xmid)/scal))
logist <- deriv(~Asym/(1+exp(-(x-xmid)/scal)),
    c("Asym", "xmid", "scal"), function(x, Asym, xmid, scal){})
Asym <- 180; xmid <- 700; scal <- 300
logist(Orange$age[1:7], Asym, xmid, scal)
```
```{r, nls-Orange}
fm1Oran.nls <- nls(circumference ~ logist(age, Asym, xmid, scal),
                   data = Orange,
                   start = c(Asym = 170, xmid = 700, scal = 500))
summary(fm1Oran.nls)
plot(fm1Oran.nls) # Fig. 8.2  (p. 341)
plot(fm1Oran.nls, Tree ~ resid(.), abline = 0)  # Figure 8.3 (p. 342)
```


###  8.1.2  Self-Starting Nonlinear Model Functions - p. 342

Bates and Watts (1988, 3.2) describe techniques for determining starting estimates in a nonlinear regression, e.g.,

- taking advantage of _partially linear_ models, so that initial estimates are needed only for those parameters that enter the model **non**linearly; in R, read docu on  `nls(*, algorithm="plinear")`.


```{r, initialValues-Orange}
Orange.sortAvg <- sortedXyData("age", "circumference", Orange)
Orange.sortAvg
NLSstClosestX(Orange.sortAvg, 130)
logistInit <- function(mCall, LHS, data) {
    xy <- sortedXyData(mCall[["x"]], LHS, data)
    if(nrow(xy) < 3)
        stop("Too few distinct input values to fit a logistic")
    Asym <- max(abs(xy[,"y"]))
    if (Asym != max(xy[,"y"])) Asym <- -Asym  # negative asymptote
    xmid <- NLSstClosestX(xy, 0.5 * Asym)
    scal <- NLSstClosestX(xy, 0.75 * Asym) - xmid
    value <- c(Asym, xmid, scal)
    names(value) <- mCall[c("Asym", "xmid", "scal")]
    value
}
## Turn it into a "self-starting" model :
logist <- selfStart(logist, initial = logistInit)
class(logist)
logist <- selfStart(~ Asym/(1 + exp(-(x - xmid)/scal)),
                    initial = logistInit,
                    parameters = c("Asym", "xmid", "scal"))
getInitial(circumference ~ logist(age, Asym, xmid, scal), Orange)
nls(circumference ~ logist(age, Asym, xmid, scal), Orange)
getInitial(circumference ~ SSlogis(age,Asym, xmid, scal), Orange)
nls(circumference ~ SSlogis(age, Asym, xmid, scal), Orange)
```

###  8.1.3  Separate Nonlinear Fits by Group: The `nlsList()` Function -  p. 347

```{r, nlsList-Orange}
fm1Oran.lis <-
   nlsList(circumference ~ SSlogis(age, Asym, xmid, scal) | Tree,
           data = Orange)
fm1Oran.lis <- nlsList(SSlogis, Orange)
fm1Oran.lis.noSS <-
    nlsList(circumference ~ Asym/(1+exp(-(age-xmid)/scal)),
             data = Orange,
             start = c(Asym = 170, xmid = 700, scal = 500))
fm1Oran.lis
summary(fm1Oran.lis)
plot(intervals(fm1Oran.lis), layout = c(3,1))
plot(fm1Oran.lis, Tree ~ resid(.), abline = 0)
```

```{r, nlsList-Theoph}
Theoph[1:4,]
fm1Theo.lis <- nlsList(conc ~ SSfol(Dose, Time, lKe, lKa, lCl),
                       data = Theoph)
fm1Theo.lis
plot(intervals(fm1Theo.lis), layout = c(3,1))
pairs(fm1Theo.lis, id = 0.1)
```


## 8.2 Fitting Nonlinear Mixed-Effects Models with nlme  (p. 354)

Again, the `Orange` growth data; no need to specify groups, as `Orange` is a `r "groupedData"`
object:

```{r, nlme-Orange, warning = FALSE}
fm1Oran.nlme <-
   nlme(circumference ~ SSlogis(age, Asym, xmid, scal),
       data = Orange,
       ## random is omitted - by default it is equal to fixed (too large here)
       fixed = Asym + xmid + scal ~ 1,
       start = fixef(fm1Oran.lis),
       ## MMä: to achieve convergence (with warnings (not shown)):
       control = nlmeControl(opt = "nlm",
                             tolerance = 1e-3, # default tolerance fails
                             pnlsMaxIter = 50,# Verbose : print.level in {0,1,2}
                             msMaxIter = 999, msVerbose = 0))

summary(fm1Oran.nlme)
# Compare with prev. "global" model
summary(fm1Oran.nls)
pairs(fm1Oran.nlme)
# set "non-significant" random effects to zero :
fm2Oran.nlme <- update(fm1Oran.nlme, random = Asym ~ 1)
anova(fm1Oran.nlme, fm2Oran.nlme) # -> not different
plot(fm1Oran.nlme) # T.A.-plot  [Figure 8.10, p.361]
## Plot "augmented" Pred.  level = 0:1  => fixed (0) & within-group (1) pred.
plot(augPred(fm2Oran.nlme, level = 0:1),  layout = c(5,1)) # Figure 8.11
qqnorm(fm2Oran.nlme, abline = c(0,1))                      # Figure 8.12
```

#### Theophylline Kinetics

Nodeling `Theoph`, now via NLME :

```{r, nlme-Theo, warning = FALSE}
(fm1Theo.nlme <- nlme(fm1Theo.lis))
intervals(fm1Theo.nlme, which="var-cov") ## ==> all corr. ~= 0
(fm2Theo.nlme <- update(fm1Theo.nlme,
                    random = pdDiag(lKe + lKa + lCl ~ 1)))
fm3Theo.nlme <-  update(fm2Theo.nlme,
                    random = pdDiag(lKa + lCl ~ 1))
anova(fm1Theo.nlme, fm3Theo.nlme, fm2Theo.nlme)
```
The simplest model (in the middle !) seems good enough:
```{r fm3Theo-plots}
plot(fm3Theo.nlme) # T.A.
qqnorm(fm3Theo.nlme, ~ ranef(.))
```

#### `CO2` Carbon Dioxide Uptake (p. 398 ff)

The $CO_2$  (grouped)Data is about _cold tolerance_ of a $C_4$ grass species.
Look at the data and compute first models, assuming the `SSasympOff()` 3-parameter exponential model:

```{r, CO2-data}
CO2
plot(CO2, outer = ~Treatment*Type, layout = c(4,1))
```
```{r, nlsL+nlme-CO2}
(fm1CO2.lis <- nlsList(SSasympOff, CO2))
(fm1CO2.nlme <- nlme(fm1CO2.lis))
(fm2CO2.nlme <- update(fm1CO2.nlme, random = Asym + lrc ~ 1))
anova(fm1CO2.nlme, fm2CO2.nlme)
plot(fm2CO2.nlme,id = 0.05,cex = 0.8,adj = -0.5)

fm2CO2.nlmeRE <- ranef(fm2CO2.nlme, augFrame = TRUE)
fm2CO2.nlmeRE
class(fm2CO2.nlmeRE)
plot(fm2CO2.nlmeRE, form = ~ Type * Treatment)
contrasts(CO2$Type)
contrasts(CO2$Treatment)
```

```{r, more-nlme-CO2}
fm3CO2.nlme <- update(fm2CO2.nlme,
        fixed = list(Asym ~ Type * Treatment, lrc + c0 ~ 1),
        start = c(32.412, 0, 0, 0, -4.5603, 49.344))
summary(fm3CO2.nlme)
anova(fm3CO2.nlme, Terms = 2:4)
fm3CO2.nlmeRE <- ranef(fm3CO2.nlme, aug = TRUE)
plot(fm3CO2.nlmeRE, form = ~ Type * Treatment)
fm3CO2.fix <- fixef(fm3CO2.nlme)
fm4CO2.nlme <- update(fm3CO2.nlme,
                fixed = list(Asym + lrc ~ Type * Treatment, c0 ~ 1),
                start = c(fm3CO2.fix[1:5], 0, 0, 0, fm3CO2.fix[6]))
summary(fm4CO2.nlme)
fm5CO2.nlme <- update(fm4CO2.nlme, random = Asym ~ 1)
anova(fm4CO2.nlme, fm5CO2.nlme)

CO2$type <- 2 * (as.integer(CO2$Type) - 1.5)
CO2$treatment <- 2 * (as.integer(CO2$Treatment) - 1.5)
fm1CO2.nls <- nls(uptake ~ SSasympOff(conc, Asym.Intercept +
  Asym.Type * type + Asym.Treatment * treatment +
  Asym.TypeTreatment * type * treatment, lrc.Intercept +
  lrc.Type * type + lrc.Treatment * treatment +
  lrc.TypeTreatment * type * treatment, c0),
  data = CO2,
  start = c(Asym.Intercept = 32.371, Asym.Type = -8.0086,
    Asym.Treatment = -4.2001, Asym.TypeTreatment = -2.7253,
    lrc.Intercept = -4.5267, lrc.Type =  0.13112,
    lrc.Treatment = 0.093928, lrc.TypeTreatment = 0.17941,
    c0 = 50.126))
anova(fm5CO2.nlme, fm1CO2.nls) ## nlme, i.e., Random effect clearly needed

## Figure 8.19, p.377 :
plot(augPred(fm5CO2.nlme, level = 0:1),  ## FIXME: problem with levels
     layout = c(6,2))  ## Actually a problem with contrasts.
```

#### Clinical Study of `Quinidine` (p.378 ff)

```{r, Quin.nlme, cache=TRUE}
fm1Quin.nlme <-
 nlme(conc ~ quinModel(Subject, time, conc, dose, interval,
                       lV, lKa, lCl),
      data = Quinidine, fixed = lV + lKa + lCl ~ 1,
      random = pdDiag(lV + lCl ~ 1), groups =  ~ Subject,
      start = list(fixed = c(5, -0.3, 2)),
      na.action = NULL, naPattern =  ~ !is.na(conc), verbose = TRUE)
fm1Quin.nlme
```

Which covariates may account for the patient-to-patient variation?
Extract random effects _augmented_ with mean/mode of covariates,
visualized in Figure 8.20 (p.382):
```{r ranef-plot-Quin.nlme}
fm1Quin.nlmeRE <- ranef(fm1Quin.nlme, aug = TRUE)
fm1Quin.nlmeRE[1:3,]
plot(fm1Quin.nlmeRE, form = lCl ~  Age + Smoke + Ethanol +
     Weight + Race + Height + glyco + Creatinine + Heart,
     control = list(cex.axis = 0.7))
```

From, this plot (and others, not shown), we _guess_ good covariates

```{r fm2Quin.nlme}
fix.fm1Q <- fixef(fm1Quin.nlme)
fm2Quin.nlme <- update(fm1Quin.nlme,
                  fixed = list(lCl ~ glyco, lKa + lV ~ 1),
                  start = c(fix.fm1Q[3], 0, fix.fm1Q[2:1]))
```

```{r fm3Quin.nlme}
fm2Quin.nlme <-
    nlme(conc ~ quinModel(Subject, time, conc, dose, interval,
                          lV, lKa, lCl),
         data = Quinidine, fixed = list(lCl ~ glyco, lV + lKa ~ 1),
         random = pdDiag(diag(c(0.3,0.3)), form = lV + lCl ~ 1),
         groups =  ~ Subject,
         start = list(fixed = c(2.5, 0, 5.4, -0.2)),
         na.action = NULL, naPattern =  ~ !is.na(conc))
summary(fm2Quin.nlme)  # wrong values
options(contrasts = c("contr.treatment", "contr.poly"))
fm2Quin.fix <- fixef(fm2Quin.nlme)
try( ## subsequent fits don't work
  fm3Quin.nlme <- update(fm2Quin.nlme,
              fixed = list(lCl ~ glyco + Creatinine, lKa + lV ~ 1),
              start = c(fm2Quin.fix[1:2], 0.2, fm2Quin.fix[3:4]))
) # .... in PNLS step
```

```{r fm4Quin.nlme, eval=FALSE, echo=FALSE }
summary(fm3Quin.nlme)
fm3Quin.fix <- fixef(fm3Quin.nlme)
fm4Quin.nlme <- update(fm3Quin.nlme,
        fixed = list(lCl ~ glyco + Creatinine + Weight, lKa + lV ~ 1),
        start = c(fm3Quin.fix[1:3], 0, fm3Quin.fix[4:5]))
##
summary(fm4Quin.nlme)
```

#### Wafer data Modeling Example

```{r, fm1Wafer, cache=TRUE}
if(FALSE) ## This fit just ping-pongs [in R]
fm1Wafer.nlmeR <-
   nlme(current ~ A + B * cos(4.5679 * voltage) +
        C * sin(4.5679 * voltage), data = Wafer,
        fixed = list(A ~ voltage + I(voltage^2), B + C ~ 1),
        random = list(Wafer = A ~ voltage + I(voltage^2),
        Site = pdBlocked(list(A~1, A ~ voltage + I(voltage^2) -1))),
        method = "REML",
        start = c(-4.255, 5.622, 1.258, -0.09555, 0.10434),
        control = nlmeControl(## default: opt = "nlmin"
            tolerance = 0.02, # < allow "rough" convergence; not sufficient
            msMaxIter = 500, msVerbose = 5)
        )

## Swapping the optimizer "saves" the situation:
fm1Wafer.nlmeR <-
   nlme(current ~ A + B * cos(4.5679 * voltage) +
        C * sin(4.5679 * voltage), data = Wafer,
        fixed = list(A ~ voltage + I(voltage^2), B + C ~ 1),
        random = list(Wafer = A ~ voltage + I(voltage^2),
        Site = pdBlocked(list(A~1, A ~ voltage + I(voltage^2) -1))),
        method = "REML",
        start = c(-4.255, 5.622, 1.258, -0.09555, 0.10434),
        control = nlmeControl(opt = "nlm",
                              tolerance = 1e-4, # default tolerance fails
                              msMaxIter = 500, msVerbose = 1)
        )

fm1Wafer.nlmeR

## ML (instead of REML) :
fm1Wafer.nlme <- update(fm1Wafer.nlmeR, method = "ML")
```

### 8.2.3  Fitting Multilevel `nlme` Models

#### Manufacturing of Analog MOS Circuits (`Wafer` data), p.386 ff

```{r, fm2Wafer.nlme, cache=TRUE}
(fm2Wafer.nlme <-
 nlme(current ~ A + B * cos(w * voltage + pi/4),
      data = Wafer,
      fixed = list(A ~ voltage + I(voltage^2), B + w ~ 1),
      random = list(Wafer = pdDiag(list(A ~ voltage + I(voltage^2), B + w ~ 1)),
      Site = pdDiag(list(A ~ voltage+I(voltage^2), B ~ 1))),
      start = c(-4.255, 5.622, 1.258, -0.09555, 4.5679)))

plot(fm2Wafer.nlme, resid(.) ~ voltage | Wafer,
     panel = function(x, y, ...) {
         panel.grid()
         panel.xyplot(x, y)
         panel.loess(x, y, lty = 2)
         panel.abline(0, 0)
     })
anova(fm1Wafer.nlme, fm2Wafer.nlme, test = FALSE)
if(FALSE) # Fails w  "Non-pos.def. approx. variance-covariance"
intervals(fm2Wafer.nlme)
```

## 8.3  Extending the Basic nlme Model  (p. 391)

Instead of i.i.d $eps_{ij} ~\sim \N(0, \sigma^2)$,
using _models_ for the variance-covariance matrix of $\eps_i$ which allows modeling
of the variance and of the covariances.

The Theophyll currently all fail  (using `eval = FALSE` knitr option):
```{r Theo-varC.Power, eval=FALSE}
fm4Theo.nlme <- update(fm3Theo.nlme,
        weights = varConstPower(power = 0.1),
        control = nlmeControl(tolerance = 1e-3)
)
```
this fit is way off

```{r, eval=FALSE, echo=FALSE}
#fm4Theo.nlme
#anova(fm3Theo.nlme, fm4Theo.nlme)
#plot(fm4Theo.nlme)
## xlim used to hide an unusually high fitted value and enhance
## visualization of the heteroscedastic pattern
plot(fm4Quin.nlme, xlim = c(0, 6.2))
```

```{r, eval=FALSE, echo=FALSE}
## fm5Quin.nlme <- update(fm4Quin.nlme, weights = varPower())
## summary(fm5Quin.nlme)
## anova(fm4Quin.nlme, fm5Quin.nlme)
## plot(fm5Quin.nlme, xlim = c(0, 6.2))
```

### 8.3.2  Correlation Structures in `nlme`,  p.395 ff

#### Counts of Ovarian Follicles - `Ovary`

Subject-specific periodic function of time (chapter 5) -- here with subject-specific frequency `w` :

```{r, nlme-Ovary}
fm1Ovar.nlme <- nlme(follicles ~ A + B * sin(2 * pi * w * Time) +
                         C * cos(2 * pi * w *Time),
                     data = Ovary,
                     fixed = A + B + C + w ~ 1, random = pdDiag(A + B + w ~ 1),
                     start = c(12.18, -3.298, -0.862, 1))
                  #  start = c(fixef(fm5Ovar.lme), 1)
fm1Ovar.nlme
ACF(fm1Ovar.nlme)
plot(ACF(fm1Ovar.nlme,  maxLag = 10), alpha = 0.05) #- Fig. 8.27
```
Hence, assume and model autocorrelation of errors:
```{r, nlme-AR-Ovary}
fm2Ovar.nlme <- update(fm1Ovar.nlme, correlation = corAR1(0.311))
fm3Ovar.nlme <- update(fm1Ovar.nlme, correlation = corARMA(p=0, q=2))
anova(fm1Ovar.nlme, fm2Ovar.nlme, fm3Ovar.nlme, test = FALSE)
if(FALSE) # "Non-positive definite .. var.-covariance"
intervals(fm2Ovar.nlme)
fm4Ovar.nlme <- update(fm2Ovar.nlme, random = A ~ 1)
anova(fm2Ovar.nlme, fm4Ovar.nlme)
##if (interactive())
```

An "intermediate" model between AR(1) and MA(2) : ARMA(1,1)
```{r, nlme-ARMA+ML-Ovary}
fm5Ovar.nlme <- update(fm4Ovar.nlme, correlation = corARMA(p=1, q=1))
anova(fm4Ovar.nlme, fm5Ovar.nlme)
plot(ACF(fm5Ovar.nlme,  maxLag = 10, resType = "n"),
       alpha = 0.05)
```

Now, it is illustrative to compare then `lnme` fit with the `lme` fit
corresponding to `fm5Ovar.lme` (of 5.3.4 --> ./ch05.Rmd )
```{r, load-ch05-fmOvar, echo=FALSE}
fm5Ovar.lme <- readRDS("fm5Ovar_lme.rds")
```
```{r, compare-lme+ML-Ovary}
fm5Ovar.lmeML <- update(fm5Ovar.lme, method = "ML")
intervals(fm5Ovar.lmeML)
fm6Ovar.lmeML <- update(fm5Ovar.lmeML, random = ~1)
anova(fm5Ovar.lmeML, fm6Ovar.lmeML)
anova(fm6Ovar.lmeML, fm5Ovar.nlme)
intervals(fm5Ovar.nlme, which = "fixed")
```


###  8.3.3 Fitting Extended Nonlin. Regr. Models w/ `gnls()`,  p.401 ff

#### High-Flux Hemodialyzer Ultrafiltration Rates - `Dialyzer`

```{r Dialyzer-ex}
fm1Dial.lis <-
  nlsList(rate ~ SSasympOff(pressure, Asym, lrc, c0) | QB,
           data = Dialyzer)
fm1Dial.lis
plot(intervals(fm1Dial.lis))
fm1Dial.gnls <- gnls(rate ~ SSasympOff(pressure, Asym, lrc, c0),
  data = Dialyzer, params = list(Asym + lrc ~ QB, c0 ~ 1),
  start = c(53.6, 8.6, 0.51, -0.26, 0.225))
fm1Dial.gnls
Dialyzer$QBcontr <- 2 * (Dialyzer$QB == 300) - 1
fm1Dial.nls <-
  nls(rate ~ SSasympOff(pressure, Asym.Int + Asym.QB * QBcontr,
  lrc.Int + lrc.QB * QBcontr, c0), data = Dialyzer,
  start = c(Asym.Int = 53.6, Asym.QB = 8.6, lrc.Int = 0.51,
  lrc.QB = -0.26, c0 = 0.225))
summary(fm1Dial.nls)
logLik(fm1Dial.nls)
plot(fm1Dial.gnls, resid(.) ~ pressure, abline = 0)
fm2Dial.gnls <- update(fm1Dial.gnls,
                       weights = varPower(form = ~ pressure))
anova(fm1Dial.gnls, fm2Dial.gnls)
ACF(fm2Dial.gnls, form = ~ 1 | Subject)
plot(ACF(fm2Dial.gnls, form = ~ 1 | Subject), alpha = 0.05)
fm3Dial.gnls <-
 update(fm2Dial.gnls, corr = corAR1(0.716, form = ~ 1 | Subject))
fm3Dial.gnls
intervals(fm3Dial.gnls)
anova(fm2Dial.gnls, fm3Dial.gnls)
```

restore two fitted models

```{r }
fm2Dial.lme <-
  lme(rate ~(pressure + I(pressure^2) + I(pressure^3) + I(pressure^4))*QB,
      Dialyzer, ~ pressure + I(pressure^2),
      weights = varPower(form = ~ pressure))
fm2Dial.lmeML <- update(fm2Dial.lme, method = "ML")
fm3Dial.gls <-
  gls(rate ~(pressure + I(pressure^2) + I(pressure^3) + I(pressure^4))*QB,
      Dialyzer, weights = varPower(form = ~ pressure),
      corr = corAR1(0.771, form = ~ 1 | Subject))
fm3Dial.glsML <- update(fm3Dial.gls, method = "ML")
anova( fm2Dial.lmeML, fm3Dial.glsML, fm3Dial.gnls, test = FALSE)
```

cleanup

```{r }
proc.time()
```

